---
title: "OmicsKit Full Tutorial – DEA, Pathway Analysis & Pathway Clustering"
author: "OmicsKit Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{OmicsKit Full Tutorial – DEA, Pathway Analysis & Pathway Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  warning   = FALSE,
  message   = FALSE,
  eval      = TRUE,
  fig.width = 7,
  fig.height = 5
)
```

# Overview

This tutorial covers the three main analytical workflows implemented in **OmicsKit**:

1. **Differential Expression Analysis (DEA) Workflow** – from raw counts to visualisation.
2. **Pathway Analysis Workflow** – GSEA result processing and plotting.
3. **Pathway Clustering Workflow** – leading-edge heatmaps and gene-set clustering.

All examples use the TCGA-LUAD datasets (`raw_counts` and `sampledata`) shipped with the package.  
Functions that require external files (e.g. `merge_GSEA`, `heatmap_GSEA`) are wrapped in optional,
clearly-labelled code blocks.

---

# 1 – Load package and data

```{r load-package}
library(OmicsKit)

# Package datasets
data("raw_counts")   # integer matrix: genes × samples
data("sampledata")   # data frame: sample metadata

# Quick look
dim(raw_counts)
head(sampledata[, 1:5])
```

---

# 2 – DEA Workflow

## 2.1 Power Analysis

Before committing resources to an experiment it is important to estimate the minimum
number of replicates needed to detect a biologically meaningful effect with sufficient
confidence.  `power_analysis()` implements the analytical framework described by
Hart *et al.* (2013, *RNA* **20**: 1684–1696).

**Statistical background (Hart *et al.* 2013):**

The approach models RNA-seq counts with a Negative Binomial distribution whose
variance is parameterised by a *dispersion* coefficient (biological coefficient of
variation squared, σ²). For a two-group comparison (treatment vs. control), the
standard error of the log₂-fold-change estimator is:

$$SE = \sqrt{\frac{2\,\sigma^2}{n}}$$

where *n* is the number of replicates per group. Under this model the non-centrality
parameter that governs statistical power is:

$$z_{\text{effect}} = \frac{|\Delta|}{SE} = \frac{|\Delta|}{\sqrt{2\,\sigma^2 / n}}$$

Because thousands of genes are tested simultaneously, a multiple-testing correction
must be applied. Hart *et al.* advocate controlling the **False Discovery Rate (FDR)**
via a Bonferroni-like adjustment on the null genes:

$$z_\alpha = \Phi^{-1}\!\left(1 - \frac{\alpha}{2\,m_0}\right)$$

where *m*₀ is the number of truly non-differentially expressed genes and Φ⁻¹ is the
normal quantile function. The realised power at sample size *n* is then:

$$\pi(n) = \Phi\!\left(z_{\text{effect}} - z_\alpha\right) + \Phi\!\left(-z_{\text{effect}} - z_\alpha\right)$$

`power_analysis()` sweeps *n* from 2 to `max_n`, computes π(*n*) at each point, and
reports the minimum *n* that meets the `power_target`.  The red dashed line marks the
desired power; the blue dashed line marks the minimum sufficient sample size.

```{r power-analysis}
?power_analysis

pa <- power_analysis(
  effect_size  = 1,       # log2 fold-change to detect
  dispersion   = 0.1,     # biological variance (σ²)
  n_genes      = 20000,   # total genes tested
  prop_de      = 0.05,    # expected proportion of DE genes
  alpha        = 0.05,    # desired FDR
  power_target = 0.8,     # desired power (1 – β)
  max_n        = 20,      # max sample size to explore
  plot         = TRUE
)

cat("Minimum sample size per group:", pa$min_sample_size, "\n")
```

```{r validate-pa}
stopifnot(exists("pa"))
stopifnot(is.list(pa))
stopifnot("min_sample_size" %in% names(pa))
stopifnot("power_table"     %in% names(pa))
stopifnot(is.numeric(pa$min_sample_size))
```

## 2.2 Get Annotations

`get_annotations()` queries Ensembl BioMart to retrieve gene metadata (symbol,
biotype, chromosome, length) for a vector of Ensembl IDs.  It requires internet
access and the `biomaRt` package.

```{r get-annotations, eval=FALSE}
?get_annotations

# OPTIONAL – requires internet access and the biomaRt package
# annotations <- get_annotations(
#   ensembl_ids = rownames(raw_counts),
#   mode        = "genes",
#   version     = "Current",
#   filename    = "gene_annotations",
#   format      = "csv"
# )
```

For the remainder of this tutorial we use a minimal annotations stub so that all
downstream chunks run without internet access.

```{r annotations-stub}
# Minimal stub: symbol and gene_length columns required by tpm() and add_annotations()
set.seed(42)
annotations_stub <- data.frame(
  geneID      = rownames(raw_counts),
  symbol      = paste0("Gen_", seq_len(nrow(raw_counts))),
  gene_length = sample(500:5000, nrow(raw_counts), replace = TRUE),
  stringsAsFactors = FALSE
)
rownames(annotations_stub) <- annotations_stub$geneID
```

## 2.3 Add Annotations

`add_annotations()` left-joins annotation columns from a reference table into a
counts matrix using the Ensembl ID as the key.

```{r add-annotations}
?add_annotations

# Add symbol and length to the counts matrix
counts_ann <- add_annotations(
  object     = raw_counts,
  reference  = annotations_stub,
  variables  = c("symbol", "gene_length"),
  data_frame = TRUE
)

head(counts_ann[, c("geneID", "symbol", "gene_length", colnames(raw_counts)[1:2])])
```

```{r validate-ann}
stopifnot(exists("counts_ann"))
stopifnot("symbol"      %in% colnames(counts_ann))
stopifnot("gene_length" %in% colnames(counts_ann))
```

## 2.4 TPM Normalisation

TPM (Transcripts Per Million) removes library-size and gene-length biases,
making expression values comparable both within and across samples.

```{r tpm-norm}
?tpm

tpm_matrix <- tpm(
  raw_counts   = raw_counts,
  gene_lengths = annotations_stub$gene_length
)

# Verify column sums ≈ 1 × 10⁶
round(colSums(tpm_matrix)[1:3])
```

```{r validate-tpm}
stopifnot(exists("tpm_matrix"))
stopifnot(nrow(tpm_matrix) == nrow(raw_counts))
stopifnot(ncol(tpm_matrix) == ncol(raw_counts))
# Each column should sum to ~1e6
stopifnot(all(abs(colSums(tpm_matrix) - 1e6) < 1))
```

## 2.5 Detect & Filter Expressed Genes

`detect_filter()` identifies genes that are reliably detectable: a gene is retained
if its **baseMean** (average normalised count across all samples) and its **group mean**
(mean in the more-expressed group) both exceed a user-defined threshold, *and* its
|log₂FC| is above a minimum fold-change floor.  This pre-filtering step removes lowly
expressed genes that would otherwise inflate the multiple-testing burden and decrease
power.  The function can handle up to three pairwise comparisons simultaneously and
returns a named list containing the filtered result data frames and the union of
detectable gene IDs.

```{r detect-filter}
?detect_filter

# Build a minimal DESeq2-like result data frame from TPM values
# (in practice this comes from DESeq2::results())
tumor_ids  <- sampledata$patient_id[sampledata$sample_type == "tumor"]
normal_ids <- sampledata$patient_id[sampledata$sample_type == "normal"]

# Ensure column name alignment
tumor_ids  <- tumor_ids[tumor_ids  %in% colnames(tpm_matrix)]
normal_ids <- normal_ids[normal_ids %in% colnames(tpm_matrix)]

mean_tumor  <- rowMeans(tpm_matrix[, tumor_ids,  drop = FALSE])
mean_normal <- rowMeans(tpm_matrix[, normal_ids, drop = FALSE])

# Simulate a DEA result data frame
set.seed(1)
deseq_sim <- data.frame(
  ensembl        = rownames(tpm_matrix),
  baseMean       = rowMeans(tpm_matrix),
  log2FoldChange = log2((mean_tumor + 1) / (mean_normal + 1)),
  pvalue         = runif(nrow(tpm_matrix)),
  padj           = p.adjust(runif(nrow(tpm_matrix)), method = "BH"),
  stringsAsFactors = FALSE
)
rownames(deseq_sim) <- deseq_sim$ensembl

# Detectable genes: baseMean > 50, group mean > 50, |L2FC| > 0
detectable <- detect_filter(
  norm.counts       = as.data.frame(tpm_matrix),
  df.BvsA           = deseq_sim,
  samples.baseline  = normal_ids,
  samples.condition1 = tumor_ids,
  cutoffs           = c(50, 50, 0)
)

cat("Detectable genes:", length(detectable$DetectGenes), "\n")
```

```{r validate-detect}
stopifnot(exists("detectable"))
stopifnot("DetectGenes"  %in% names(detectable))
stopifnot("Comparison1"  %in% names(detectable))
stopifnot(is.character(detectable$DetectGenes))
```

## 2.6 Dimensionality Reduction

### PCA, t-SNE and UMAP – conceptual differences

| Method | Core idea | Preserves | When to use |
|--------|-----------|-----------|-------------|
| **PCA** | Linear rotation that maximises variance explained by successive orthogonal axes | Global structure, variance | First pass; interpretable axes (PC1 captures most variance) |
| **t-SNE** | Minimises KL-divergence between high- and low-dimensional neighbourhood distributions | Local structure | Revealing fine clusters; axes have no direct biological meaning; sensitive to `perplexity` |
| **UMAP** | Graph-based manifold learning (fuzzy topological structure) | Both local *and* some global structure | Faster than t-SNE on large datasets; generally preferred when both local clusters and inter-cluster relationships matter |

### 2.6.1 PCA

```{r nice-pca}
?nice_PCA

# Prepare sample-level annotation (id column required)
annot_pca <- data.frame(
  id          = colnames(tpm_matrix),
  sample_type = sampledata$sample_type[match(colnames(tpm_matrix), sampledata$patient_id)],
  stringsAsFactors = FALSE
)

nice_PCA(
  object       = tpm_matrix,
  annotations  = annot_pca,
  PCs          = c(1, 2),
  ntop         = 500,
  variables    = c(fill = "sample_type"),
  legend_names = c(fill = "Sample Type"),
  colors       = c("tumor" = "#E74C3C", "normal" = "#2ECC71"),
  shapes       = 21,
  size         = 5,
  title        = "PCA – TCGA-LUAD",
  transform    = TRUE
)
```

### 2.6.2 t-SNE

```{r nice-tsne}
?nice_tSNE

nice_tSNE(
  object       = tpm_matrix,
  annotations  = annot_pca,
  perplexity   = 3,
  max_iterations = 500,
  seed         = 42,
  variables    = c(fill = "sample_type"),
  legend_names = c(fill = "Sample Type"),
  colors       = c("tumor" = "#E74C3C", "normal" = "#2ECC71"),
  shapes       = 21,
  size         = 5,
  title        = "t-SNE – TCGA-LUAD",
  transform    = TRUE
)
```

### 2.6.3 UMAP

```{r nice-umap}
?nice_UMAP

nice_UMAP(
  object       = tpm_matrix,
  annotations  = annot_pca,
  neighbors    = 5,
  components   = 2,
  epochs       = 200,
  seed         = 42,
  variables    = c(fill = "sample_type"),
  legend_names = c(fill = "Sample Type"),
  colors       = c("tumor" = "#E74C3C", "normal" = "#2ECC71"),
  shapes       = 21,
  size         = 5,
  title        = "UMAP – TCGA-LUAD",
  transform    = TRUE
)
```

## 2.7 Violin-Scatter-Box (VSB) Plot

`nice_VSB()` overlays individual data points on a violin+box plot for a single
gene — ideal for communicating expression distributions across groups.

```{r nice-vsb}
?nice_VSB

# Build a minimal VST-like matrix (log2-TPM here for demonstration)
vst_matrix <- log2(tpm_matrix + 1)

# Use only samples with known sample_type
annot_vsb <- data.frame(
  id          = colnames(vst_matrix),
  sample_type = sampledata$sample_type[match(colnames(vst_matrix), sampledata$patient_id)],
  stringsAsFactors = FALSE
)
annot_vsb <- annot_vsb[!is.na(annot_vsb$sample_type), ]
vst_sub   <- vst_matrix[, annot_vsb$id, drop = FALSE]

# Pick a gene that is present
gene_of_interest <- rownames(vst_sub)[1]

nice_VSB(
  object     = vst_sub,
  annotations= annot_vsb,
  variables  = c(fill = "sample_type"),
  genename   = gene_of_interest,
  symbol     = "",
  labels     = c("N", "T"),
  categories = c("normal", "tumor"),
  colors     = c("normal" = "#2ECC71", "tumor" = "#E74C3C"),
  shapes     = 21,
  markersize = 3,
  alpha      = 0.8,
  jitter     = 0.15
)
```

## 2.8 Volcano Plot

The Volcano plot simultaneously displays fold-change (x-axis) and statistical
significance (y-axis, −log₁₀ FDR), making it easy to identify genes with both
large and significant changes.

```{r nice-volcano}
?nice_Volcano

# Add gene symbols for labelling
deseq_sim$symbol <- annotations_stub$symbol[
  match(deseq_sim$ensembl, annotations_stub$geneID)
]
# Replace NA padj with 1
deseq_sim$padj[is.na(deseq_sim$padj)] <- 1

nice_Volcano(
  results   = deseq_sim,
  x_var     = "log2FoldChange",
  y_var     = "padj",
  label_var = "symbol",
  cutoff_x  = 1,
  cutoff_y  = 0.05,
  x_range   = 6,
  y_max     = 6,
  title     = "Volcano – Tumor vs Normal",
  colors    = c("red", "grey70", "blue"),
  legend    = TRUE
)
```

## 2.9 Save Results

`save_results()` writes three Excel files from a DEA result data frame: all genes,
up-regulated, and down-regulated genes.

```{r save-results, eval=FALSE}
?save_results

# OPTIONAL – writes Excel files to the working directory
# save_results(
#   df           = deseq_sim,
#   name         = "TumorVsNormal",
#   l2fc         = 1,
#   cutoff_alpha = 0.05
# )
```

## 2.10 Split Cases

`split_cases()` partitions genes from three pairwise comparisons into up to ten
mutually exclusive expression-pattern categories.

```{r split-cases, eval=FALSE}
?split_cases

# OPTIONAL – requires three DESeq2 result data frames
# cases <- split_cases(
#   df.BvsA              = res_TumorVsNormal,
#   df.CvsA              = res_MetastasisVsNormal,
#   df.BvsC              = res_TumorVsMetastasis,
#   unique_id            = "ensembl",
#   significance_var     = "padj",
#   significance_cutoff  = 0.05,
#   change_var           = "log2FoldChange",
#   change_cutoff        = 1
# )
```

## 2.11 Kaplan-Meier Survival Plot

`nice_KM()` fits and plots Kaplan-Meier survival curves stratified by a binary
gene-expression variable.

```{r nice-km, eval=FALSE}
?nice_KM

# OPTIONAL – requires survival and survminer packages
# km_data <- data.frame(
#   TP53_muts  = c(rep("Yes", 15), rep("No", 17)),
#   OS_time    = c(sample(100:2000, 15), sample(500:3000, 17)),
#   OS_event   = sample(0:1, 32, replace = TRUE)
# )
# nice_KM(
#   data      = km_data,
#   gene      = "TP53_muts",
#   time_var  = "OS_time",
#   event_var = "OS_event"
# )
```

---

# 3 – Pathway Analysis Workflow

## 3.1 Merge GSEA Output Files

After running GSEA (e.g., via `GSEA_all.sh`), `merge_GSEA()` reads all `.tsv` result
files in a directory and combines them into a single tidy data frame.

```{r merge-gsea, eval=FALSE}
?merge_GSEA

# OPTIONAL – requires a directory with GSEA .tsv result files
# gsea_merged <- merge_GSEA(
#   input_directory = "path/to/gsea_results/",
#   output_file     = "collections_merged_gsea_data.tsv"
# )
```

## 3.2 GSEA Bar Plot

`barplot_GSEA()` creates a faceted bar chart of NES values coloured by −log₁₀(FDR).

```{r barplot-gsea-demo}
?barplot_GSEA

# Simulate a minimal GSEA-like data frame
gsea_demo <- data.frame(
  New_name    = rep(c("Hallmarks", "Reactome"), each = 6),
  datatype    = rep(c("TumorVsNormal", "MetVsNormal"), 6),
  NES         = c(1.8, -1.5, 2.1, -1.2, 1.4, -1.9, 1.6, -1.3, 2.0, -1.7, 1.1, -2.0),
  `-Log10FDR` = runif(12, 1, 5),
  stringsAsFactors = FALSE
)

custom_lbl <- c(
  TumorVsNormal = expression("Tumor vs Normal"),
  MetVsNormal   = expression("Met vs Normal")
)

barplot_GSEA(
  data          = gsea_demo,
  output_path   = tempfile(fileext = ".svg"),
  custom_labels = custom_lbl,
  axis_y        = "NES"
)
```

## 3.3 Global GSEA Plot (`plot_GSEA`)

`plot_GSEA()` generates either a faceted multi-comparison bar chart or a single-comparison
panel layout with pathway labels and MSigDB collection annotations.

```{r plot-gsea-demo}
?plot_GSEA

# Multi-comparison mode (Comparison column has >1 unique value)
gsea_multi <- data.frame(
  Comparison  = rep(c("TumorVsNormal", "MetVsNormal"), 6),
  New_name    = rep(c("Hallmarks", "Reactome", "GO"), 4),
  NES         = c(1.8, -1.5, 2.1, -1.2, 1.4, -1.9, 1.6, -1.3, 2.0, -1.7, 1.1, -2.0),
  FDR         = runif(12, 0.001, 0.05),
  stringsAsFactors = FALSE
)

plot_GSEA(
  data        = gsea_multi,
  Comparison  = "Comparison",
  axis_y      = "NES",
  fdr_col     = "FDR",
  logFDR      = TRUE,
  order       = "desc",
  ncol_wrap   = 2,
  free_y      = TRUE
)
```

---

# 4 – Pathway Clustering Workflow

## 4.1 Heatmap of Leading-Edge Genes

`heatmap_GSEA()` reads expression data, metadata, GMT gene-set definitions, and GSEA
output to generate per-pathway leading-edge heatmaps.  This function writes PDF and
JPEG files to disk and therefore requires local input files.

```{r heatmap-gsea, eval=FALSE}
?heatmap_GSEA

# OPTIONAL – requires local GSEA output files
# heatmap_GSEA(
#   main_dir         = "path/to/project/",
#   expression_file  = "expression_matrix.txt",
#   metadata_file    = "metadata.xlsx",
#   gmt_file         = "gene_sets.gmt",
#   ranked_genes_file= "ranked_genes.rnk",
#   gsea_file        = "gsea_results.tsv",
#   output_dir       = "leading_edge_heatmaps",
#   sample_col       = "Sample",
#   group_col        = "group",
#   save_dataframe   = TRUE
# )
```

---

# 5 – Validation Summary

The block below re-checks the key objects produced by the runnable chunks.

```{r validation-summary}
# Core objects from the DEA workflow
stopifnot(exists("raw_counts"))
stopifnot(exists("sampledata"))
stopifnot(exists("annotations_stub"))
stopifnot(exists("tpm_matrix"))
stopifnot(exists("detectable"))
stopifnot(exists("deseq_sim"))
stopifnot(exists("pa"))

# Structural checks
stopifnot(is.data.frame(sampledata))
stopifnot(is.numeric(tpm_matrix))
stopifnot(all(abs(colSums(tpm_matrix) - 1e6) < 1))
stopifnot("DetectGenes" %in% names(detectable))
stopifnot(is.numeric(pa$min_sample_size))

message("All validation checks passed.")
```

---

# Session Information

```{r session-info}
sessionInfo()
```
